Missing data: Current practice in football research and recommendations for reporting missingness

### 2020-09-03 paper/paper.Rmd

Missing data can undermine the validity and accuracy of research results
[@Sainani2015; @Nakagawa2008; @Sterne2009]. Data can go
missing many ways. For example, accidentally skipping a survey question,
equipment failure, or intentionally not recording values. Before
analysis, missing data must be handled, with values typically deleted or
imputed. There is no universal approach to handling missing data. 
Contextual factors--such as, the study design and objective, and
pattern of missingness--determine how missings should be handled, on a
case-by-case basis [@Sainani2015].
The effect of missing data on the validity and accuracy of a study's
results is largely determined by the reasons for missingness, and how
missing data are handled.

Data does not go missing the same way every time. There are three broad
categorisations describing why data can be missing: Missing Completely
at Random (MCAR), Missing at Random (MAR), and Missing Not at Random
(MNAR). These categorisations help describe whether the missingness
occurs by chance (MCAR), due to some observed variable (MAR) or an
unobserved variable (MNAR), with MCAR being benign and the latter two
biased. Football relevant examples are provided in Supplement 1, and
more detailed explanations of MCAR, MAR, MNAR can be found elsewhere
[@Sainani2015; @Nakagawa2008; Sterne2009].

The three categorisations of missingness types provide a useful
framework to describe bias arising from missing data. If one strongly
suspects data is MCAR, one can be more certain your data isn't biased,
otherwise there may be bias in your sample. This does not mean that data MAR
or MNAR make data invalid, instead it identifies potential biases, which
can help improve future research design **(Nick, can we look at this sentence please)**. However, these categorisations
do not necessarily guide a researcher in their subsequent actions in
analysis, or describe specifically how values go missing. To identify
possible mechanisms for missingness, the data must be explored. It can
be challenging to identify the missingness mechanism in data, but it is
critical, since missing values could change the outcome of a study. This
could mean the work cannot be reproduced, and worse, may lead scientists
and practitioners to the wrong outcome.

It is imperative that missing data, or lack thereof, are reported. However, there is 
evidence that outside of sports science, researchers seldom consider missing data [@Schafer2002]. 
While there has been significant interest in modelling missing
data [@refs], the exploration and reporting of missings have received
less attention [@Schafer2002]. This article aimed to: (1) estimate the proportion
of articles that report missing data in football research; (2) highlight
several improvable missing data practices; and (3) provide recommendations for
exploring and reporting missingness.

@@

To estimate the proportion of articles that report missing data in
football research, we conducted a systematic search [@Schafer2002] **(should this reference be here?)**. Football related
articles published in 2019 were studied. Three major sports medicine
databases (SPORTDiscus, Embase and Cinahl) were searched, using key
terms and search limits (Supplement 2). Included articles were: (a) written in English; (b) had an accessible
full-text; and (c) included quantitative data. The search process is
summarised in the PRISMA figure provided in Supplement 2.

>>>>

Missing data are values that should have been observed, but were not. Since the
values are unobserved, this can mean the population of interest might not be 
properly sampled, inducing bias into study results. Left undetected or poorly handled, missing data can undermine the validity and accuracy of research results
[@Sainani2015; @Nakagawa2008; @Sterne2009]. Data can go
missing many ways. For example, accidentally skipping a survey question,
equipment failure, or intentionally not recording values. Before
analysis, missing data must be handled, with values typically deleted or
imputed. There is no universal approach to handling missing data. 
Contextual factors--such as, the study design and objective, and
pattern of missingness--determine how missings should be handled, on a
case-by-case basis [@Sainani2015].
The effect of missing data on the validity and accuracy of a study's
results is largely determined by the reasons for missingness, and how
missing data are handled.

A missing value should have been observed or measured in the data, but was not. For example, a goal scored in football takes place, but the scorekeeper wasn't paying attention and did not record an event. This "goal" is now a missing value. Compare this to a game being rained out and no game taking place. There is no missing data because there is no data to be observed. A final example - a study screens out individuals who do not meet criteria for the study. These are not missing values, because there was no intent for these to be observed or measured.

Data does not go missing the same way every time. There are three broad
categorisations describing why data can be missing: Missing Completely
at Random (MCAR), Missing at Random (MAR), and Missing Not at Random
(MNAR). These categorisations help describe whether the missingness
occurs by chance (MCAR), due to some observed variable (MAR) or an
unobserved variable (MNAR), with MCAR being the least harmful, and MAR and MNAR
biased [@Newman2014]. Football relevant examples are provided in Supplement X, and
more detailed explanations of MCAR, MAR, MNAR can be found elsewhere
[@Sainani2015; @Nakagawa2008; Sterne2009].

The three categorisations of missingness types provide a useful
framework to describe bias arising from missing data. If you strongly
suspect data is MCAR, your results are less likely to be biased. If you suspect
MAR or MNAR, there is bias in your results. This does not mean data MAR
or MNAR make results invalid, instead it identifies potential bias, which can be
used to help improve future research design. However, these categorisations
do not necessarily guide the user in their subsequent actions in
analysis, or describe specifically how values go missing. To identify
possible mechanisms for missingness, the data must be explored. It can
be challenging to identify the missingness mechanism in data, but it is
critical, since missing values could change the outcome of a study. This
could mean the work cannot be reproduced, and worse, may lead scientists
and practitioners to the wrong outcome.

It is imperative that missing data, or lack thereof, are reported [@Schafer2002]. 
While there has been significant interest in modelling missing
data [@refs], the exploration and reporting of missings have received
less attention [@Schafer2002]. This article aimed to: (1) estimate the proportion
of articles that report missing data in football research; (2) highlight
several improvable missing data practices; and (3) provide recommendations for
exploring and reporting missingness.

@@

To estimate the proportion of articles that report missing data in
football research, we conducted a systematic search [@PRISMA]. Football related
articles published in 2019 were studied. Three major sports medicine
databases (SPORTDiscus, Embase and Cinahl) were searched, using key
terms and search limits (Supplement 2). Included articles were: (a) written in English; (b) had an accessible
full-text; and (c) included quantitative data. The search process is
summarised in the PRISMA figure provided in Supplement 2.
